Load Shifting Versus Manual Frequency Reserve: \\ Which One is More Appealing to Flexible Loads in Denmark?

### 2024-03-18 tex/review/response_letter.tex

Regarding the second item with the sensitivity analysis, we have added the following statement immediately after the above addition while also adding a plot of the sensitivity analysis for the mFRR 2021 model as an additional appendix:

\nt{The first solution strategy uses 50 scenarios which, and we found that increasing the number of scenarios further did not improve the performance as shown in Appendix B.. The same was the case for choosing five days in the second solution strategy.}

\textbf{Comment 3:} \textit{Please add text annotations to the legend to improve the legibility of Fig.7 and Fig. 8. Additionally, does the "measurements" mean the real measurements in Fig. 7 and Fig. 8? Why is there a big difference between the curve of measurements and the curve of simulated results? Is this phenomenon due to inaccurate modeling or unreasonable assumptions? Could you narrow the difference?}

\auth Thanks for the comment. We have made changes to figures 7 and 8, accordingly, by adding descriptive names in parenthesis to the legends. For example: $P_{h}^{\text{Base}}$ \textit{(Baseline power)}. The measurements are real-life data as shown in Figure 3. The point of the measurements in Figure 3 is to show how our state-space model is able to simulate reality, i.e., the measurements, to a satisfactory degree. In Section 2.1.3 (\textit{Model validation}), we explain how our state-space model fits the measurements, and how it is able to simulate freezer temperatures for a whole day. To achieve a better performance, i.e., a narrower difference to the measurements, requires even more sophisticated modelling techniques and is perhaps difficult to achieve given the current data provided. For example, there are a lot of uncertain events like people opening the lids to the freezers causing temperature drops. The important element of the state-space models used here is that they capture the most fundamental physics and dynamics from which one can directly use within an optimization model. However, we concede that it is confusing to also show the measurements in Figures 7 and 8. We have therefore removed them from those figures.

@@

\nt{Load Shifting Versus Manual Frequency Reserve:  Which One is More Appealing to Thermostatically Controlled Loads in Denmark?}

\textbf{Comment 2:} \textit{Abstract: The use of a supermarket freezer as a representative flexible load is ok. But, trying to generalize the results for all types of flexible loads, as suggested by the title, is not ok. As mentioned above, flexible loads can include many types of loads. None of the other loads were studied in this paper. I strongly suggest replacing flexible loads with thermostatically controlled loads.}

\auth Thank you for your comment. As before, we agree and we will make sure to replace all instances of \textit{flexible load} with \textit{thermostatically controlled load}

\changes The manuscript has been changed everywhere where \textit{flexible load} occurs which has been replaced with \nt{thermostatically controlled load} instead for the first instance, and \nt{TCL} for all subsequent occurrences. This includes for the whole manuscript, including the abstract and all sections.

### 2024-03-18 tex/figures/timeline_mfrr_variables_tikz.tex
Revision

  \node[align=center, draw=none, fill=none, rotate=90] at (-0.5,4) {mFRR market};
  \node[align=center, draw=none, fill=none, rotate=90] at (-0.5,0) {Load shifting};

  \draw[rounded corners, fill=RectangleColor, very thick] (0, 3) rectangle (4, 5) node[midway]{\Large mFRR \\ \Large bidding};
  \node at (2,2) {\Large  $p_{h}^{\rm{r},\uparrow}$};
  \draw[dashed, black, line width=1.5pt] (4.5,-1) -- (4.5,5);
  \node at (4.5,-2) {\Large D$-$1};
  \node at (4.5,-2.7) {\large (9:30am)};
  \node[align=center, draw=none, fill=none, rotate=90] at (4.8,1.2) {mFRR gate closure};

  \draw[dashed, black, line width=1.5pt] (7,-1) -- (7,5);
  \node at (7,-2) {\Large D$-$1};
  \node at (7,-2.7) {\large (12:00pm)};
  \node[align=center, draw=none, fill=none, rotate=90] at (6.7,1.9) {Day-ahead market clearing};
  \draw[rounded corners, fill=RectangleColor, very thick] (7.375, 1) rectangle (11.375, 3) node[midway]{\Large Load shifting};
  \node at (9.375,0) {\Large  $p_{h}^{\text{Base}}$};

  \draw[rounded corners, fill=RectangleColor, very thick] (12, 3) rectangle (16, 5) node[midway]{\Large Regulating  \\ \Large power \\ \Large bidding};
  \node at (14,2) { \Large $\lambda_{h,\omega}^{\text{bid}}$};
  \draw[dashed, black, line width=1.5pt] (16.25,-1) -- (16.25,5);
  \node[align=center, draw=none, fill=none, rotate=90] at (16.65,1.9) {Regulation bid gate closure};
  \node at (16.25,-2) {\Large D$-$1};
  \node at (16.25,-2.7) {\large (5:00pm)};

  \draw[rounded corners, fill=RectangleColor, very thick] (17.5, 3) rectangle (21.5, 5) node[midway]{\Large Activation};
  \node at (19.5,2) {\Large $\Gamma_{h,\omega}$};
  % {\Large $\Bigl( \bm{p}^{\rm{r},\uparrow}, \bm{\lambda}_{\omega}^{\text{bid}}, \bm{p}_{\omega}^{b,\uparrow}, \bm{p}_{\omega}^{b,\downarrow}, \bm{s}_{\omega}, \bm{T}_{\omega}^{c}, \bm{T}_{\omega}^{f}, \notag \\ \quad & \quad \bm{T}_{\omega}^{c, B}, \bm{T}_{\omega}^{f,B}, \bm{\phi}_{\omega}, \bm{g}_{\omega} \Bigr)$};
  \node at (19.5,-2) {\Large D};
  \node at (19.5,-2.7) {\large (12pm -- 12am)};

>>>>

  % Legend box with LightBlue color
%\node[draw, rectangle, fill=LightBlue, rounded corners=3pt, minimum width=3cm, minimum height=0.7cm, align=center] at (13.5,6) {\textbf{Load shifting}};  
  % Legend box with RectangleColor
%\node[draw, rectangle,fill=RectangleColor, rounded corners=3pt, minimum width=3cm, minimum height=0.7cm, align=center] at (9,6) {\textbf{mFRR market}};

  \draw[rounded corners, fill=RectangleColor, very thick] (0, 3) rectangle (4, 5) node[midway]{\Large mFRR \\ \Large bidding};
  \node at (2,2) {\Large  $p_{h}^{\rm{r},\uparrow}$};
  \draw[dashed, black, line width=1.5pt] (4.5,-1) -- (4.5,5);
  \node at (4.5,-2) {\Large D$-$1};
  \node at (4.5,-2.7) {\large (9:30am)};
  \node[align=center, draw=none, fill=none, rotate=90] at (4.8,2.1) {mFRR market gate closure};

  \draw[dashed, black, line width=1.5pt] (7,-1) -- (7,5);
  \node at (7,-2) {\Large D$-$1};
  \node at (7,-2.7) {\large (12:00pm)};
  \node[align=center, draw=none, fill=none, rotate=90] at (6.7,2.5) {Day-ahead market gate closure};
  \draw[rounded corners, fill=LightBlue, very thick] (7.375, 1) rectangle (11.375, 3) node[midway]{\Large Load  \\ \Large shifting};
  \node at (9.375,0) {\Large  $p_{h}^{\text{Base}}$};

  \draw[rounded corners, fill=RectangleColor, very thick] (12, 3) rectangle (16, 5) node[midway]{\Large Regulating  \\ \Large power \\ \Large bidding};
  \node at (14,2) { \Large $\lambda_{h,\omega}^{\text{bid}}$};
  \draw[dashed, black, line width=1.5pt] (16.25,-1) -- (16.25,5);
  \node[align=center, draw=none, fill=none, rotate=90] at (16.65,3.2) {Deadline for regulating bid submission};
  \node at (16.25,-2) {\Large D$-$1};
  \node at (16.25,-2.7) {\large (5:00pm)};

  \draw[rounded corners, fill=RectangleColor, very thick] (17.5, 3) rectangle (21.5, 5) node[midway]{\Large mFRR \\ \Large activation};
  \node at (19.5,2) {\Large $\Gamma_{h,\omega}$};
  % {\Large $\Bigl( \bm{p}^{\rm{r},\uparrow}, \bm{\lambda}_{\omega}^{\text{bid}}, \bm{p}_{\omega}^{b,\uparrow}, \bm{p}_{\omega}^{b,\downarrow}, \bm{s}_{\omega}, \bm{T}_{\omega}^{c}, \bm{T}_{\omega}^{f}, \notag \\ \quad & \quad \bm{T}_{\omega}^{c, B}, \bm{T}_{\omega}^{f,B}, \bm{\phi}_{\omega}, \bm{g}_{\omega} \Bigr)$};
  \node at (19.5,-2) {\Large D};
  \node at (19.5,-2.7) {\large (12pm -- 12am)};

### 2024-03-18 tex/paper_flex_v2.tex
Revision

\begin{document}
\title{
    %A Pioneering Market Framework for Promoting Demand-side Flexibility in Denmark 
    %A New Ecosystem for Demand-side Flexibility Aggregators in Denmark 
    Load Shifting Versus Manual Frequency Reserve: \\ Which One is More Appealing \textcolor{red}{to Thermostatically Controlled Loads in Denmark}?
}
% Pioneering Market Framework for the Future Power System: the Case of Denmark
%\author{Peter Alexander Vistar Gade$^{\textrm{a}}$, Henrik Bindner, Jalal
%Kazempour}
\author[1,2]{Peter A.V. Gade\footnote{Corresponding author. Tel.: +45 24263865. \\ Email addresses: pega@dtu.dk (P.A.V. Gade), Trygve.Skjotskift@ibm.com (T. Skjøtskift), chazi@dtu.dk (C. Ziras), hwbi@dtu.dk (H.W. Bindner), jalal@dtu.dk (J. Kazempour).}}
\author[2]{Trygve Skjøtskift}
\author[1]{Charalampos Ziras}
\author[1]{Henrik W. Bindner}
\author[1]{Jalal Kazempour}
\affil[1]{Department of Wind and Energy Systems, Technical University of Denmark, Kgs. Lyngby, Denmark}
\affil[2]{IBM Client Innovation Center, Copenhagen, Denmark}
%\setcounter{Maxaffil}{0}
\renewcommand\Affilfont{\itshape\small}

@@

This paper investigates %how  thermostatically controlled loads can deliver flexibility 
demand-side flexibility provision in two  forms of manual Frequency Restoration Reserve (mFRR) services and load shifting, and explores which one is financially more appealing to \textcolor{red}{Thermostatically Controlled Loads (TCLs)} in Denmark. While mFRR is an ancillary service required for the system and being bought by the system operator, load shifting is an individual act of the \textcolor{red}{TCL} in response to the variation of hourly electricity prices.  %We discuss advantages and disadvantages of the two and their appeal from a monetary point of view.
Without loss of generalization, we consider a supermarket freezer  as a representative \textcolor{red}{TCL}, and develop a grey-box model  describing its temperature dynamics using real data from a supermarket in Denmark. Taking into account price and activation uncertainties, a stochastic mixed-integer linear program is formulated to maximize the flexibility value from the freezer. %For practical reasons, we propose a linear policy to determine regulating power bids.
%, and then linearize the mFRR activation conditions. 
%For computational ease, we develop a decomposition technique, splitting the problem to a set of smaller sub-problems, one per scenario. %An out-of-sample evaluation was done on unseen Danish market price data for 2022. 
%
%We observed that load shifting was more profitable, but had a greater impact on the air and food temperatures in the freezer as opposed to mFRR that depends on the system state and bid price for activation.
%
%solution strategies for mFRR are presented: one with a simple policy and the other one with a dynamically updated policy. The proposed model accurately captures the timeline of decisions for bidding in manual frequency reserves.
%by using McCormick relaxation.
Examined on an ex-post simulation based on  Danish spot and balancing market prices in 2022, load shifting shows to be more profitable than mFRR provision, but is also more consequential for temperature deviations in the freezer. This indicates the need for regulatory measures by the Danish  system operator to make the provision of ancillary services more attractive to  \textcolor{red}{TCL}s.

@@

Power consumers are looking to reduce costs by any means due to sky-rocketing energy and gas prices. In Denmark, supermarkets are especially exposed due to their high energy consumption, and they are actively looking into initiatives to reduce their electricity bill. One of these initiatives is the application of demand-side flexibility, whereby a supermarket can shift consumption in time when spot (day-ahead) prices are comparatively lower or when the power system needs up-regulation services. There are thousands of supermarkets in Denmark willing to provide flexibility, and their freezers are especially suited for this as they constitute a big share of their energy consumption.

@@

In \cite{petersen2012eso2} and \cite{pedersen2013direct}, a complete white-box model of a supermarket refrigeration system is presented and validated against real-life data. It is also shown how such a system can provide demand response. This work serves as a  benchmark for any grey-box model of a supermarket refrigeration system. However, such an approach is hard to scale and requires complete knowledge of every refrigeration system. In \cite{pedersen2016improving}, a second-order model is used to model the food and air temperature in a freezer in an experimental setting. It is shown how food temperature has much slower dynamics than air temperature. In \cite{hao2014aggregate}, a simple first-order virtual battery model for a thermostatically controlled load is presented. This model also constitutes the starting point for modeling temperature dynamics in this work, as it has an intuitive interpretation. A similar bucket model is introduced in \cite{petersen2013taxonomy}.

@@

\textcolor{red}{In \cite{schaperow2019simulation,chanpiwat2020using, moglen2020optimal}}, residential air condition units are modeled using up- and down-regulation blocks characterizing flexibility. The blocks are obtained from grey-box models of households \cite{siemann2013performance}. The authors show how such a block formulation can be solved using exact and stochastic dynamic programming in the context of peak shaving demand for a utility in the US. The demand-side flexibility then functions as a hedge towards extreme electricity prices. Although such a use case has not been relevant in Denmark yet, the estimation of flexibility blocks is a novel idea as it avoids having to explicitly integrate a physical model into an optimization problem. However, a tradeoff is that the curse of dimensionality quickly makes dynamic programming computationally intractable for a large portfolio of heterogeneous demand-side assets. Furthermore, it might be difficult to assure the Markov property \cite{MarkovProperty} when reformulating the problem even slightly. Flexibility blocks are also used in \cite{bobo2018offering} for an offering strategy. Here, the flexibility blocks are derived from measurements of residential appliances in the Ecogrid 2.0 project \cite{ecogrid}.

@@

In \cite{biegel2013electricity}, a portfolio of residential heat pumps is modeled using a linear first-order model. Individual units are lumped together as essentially one big, aggregated heat pump as described in \cite{biegel2013lumped}. The first-order model is a grey-box model incorporating the physics, i.e., the temperature dynamics in a household. Such an approach will be used in this work as well. The authors also consider how the portfolio of heat pumps can be used for load shifting and real-time bidding in the balancing market, but do not consider mFRR service provision. However, they assume that all the assets can be controlled continuously in a  model predictive control  setting. Temperature deviations are minimized directly in the objective function using the integrated error of total deviations. This introduces a trade-off parameter in the objective function that must be tuned to weigh the economic value versus the temperature deviation compared to the setpoint.

\textcolor{red}{Similar studies exist in the literature albeit in different settings. For example, \cite{iria2018trading} uses a two-stage stochastic optimization for prosumers to bid into tertiary reserves, i.e., mFRR. A similar approach is taken in \cite{la2021mixed} although with a focus on scalability to thousands of assets and distribution of flexibility amongst prosumers. In \cite{nitsch2021economic}, the authors show how batteries can be used to provide secondary reserves. TCLs have also been investigated for provision of primary reserves, i.e., frequency reserves, in the Nordics \cite{paridari2020flexibility}, in microgrids \cite{mendieta2020primary}, and in the Australian power grid \cite{attarha2020network}}.

\subsection{Our contribution and outline}
%
To the best of our knowledge, no work has investigated the monetary value and incentives for providing mFRR services versus load shifting, although both have been studied extensively. This work aims to fill that gap by taking a holistic view on the incentives from the flexible consumer's perspective in Denmark. Furthermore, when stating the objective functions for mFRR, most studies have assumed a simplified market structure, and neglect the fact that the aggregator and the consumer are not necessarily the same entity. We discuss the consequences of this assumption in relation to the market structure in Denmark \cite{gade2022ecosystem}, and in the context of mFRR and load shifting.

@@

The rest of the paper is organized as follows. Section \ref{sec:monetizing_flex} describes modeling a supermarket freezer as a thermostatically controlled load. Thereafter, mFRR and load shifting are described in detail and their objective functions are stated. Section \ref{sec:OptimizationModel} presents the proposed stochastic optimization problem.
%Furthermore, the two solution strategies for solving the optimization problem for mFRR are introduced. 
Section \ref{sec:results} provides  results for a case study. Section \ref{sec:conclusion} concludes the paper. Finally, \textcolor{red}{Appendix A} provides the full model formulation.

\section{Describing flexibility of a thermostatically controlled load}\label{sec:monetizing_flex}

There are several ways to monetize flexibility from a \textcolor{red}{TCL}. In this section, we focus on mFRR and load shifting. First, we describe how to mathematically model a TCL as a flexible resource. Second, we describe how to monetize the flexibility from TCLs for mFRR and load shifting, and provide the objective functions in both cases. For mFRR, the objective function includes all costs and revenues for the BRP, while for load shifting, the objective function only includes the flexible demand's perspective. This approach explicitly shows the situation where flexible demand is activated without including the BRP, as this is more realistic.

\subsection{Modeling TCL as a flexible resource}
%
TCLs are characterized by being controlled such that the temperature is kept at a specified setpoint. Examples include heat pumps, freezers, air condition units, etc. They constitute an important part of demand-side flexibility due to  inherent thermal inertia of such temperature-driven systems \cite{hao2014aggregate}.

We focus on freezers only, which are a very common type of TCLs. Specifically, we focus on a single freezer display in a Danish supermarket. Freezers are characterized by a large thermal inertia due to the frozen food, which makes them suitable for flexibility. On the other hand, there is a risk of food degradation when utilizing flexibility. Therefore, it is important to model the temperature dynamics in the freezer for a realistic and risk-aware estimation of its flexibility.

@@

\begin{figure}[b]
    \centering
    \includestandalone[width=\columnwidth]{figures/timeline_mfrr_variables_tikz}
    \caption{\textcolor{red}{Variables related to the timeline of mFRR bidding decisions and load shifting.}}
    \label{fig:timeline_mfrr_variables}
\end{figure}

@@

\subsection{Load shifting}
%
Another option for utilizing flexibility is to shift the load to a different time according to the spot market prices which are known already 12-36 hours in advance \textcolor{red}{as shown in Figure \ref{fig:timeline_mfrr_variables}}. We note that in Europe and Denmark, day-ahead prices are directly exposed to consumers with no time-of-use price in between as is often the practice in North America. Then, it is simply a matter of consuming in low-price hours.
%
For a TCL, there are additional constraints to how energy can be shifted due to the rebound effect. First, there can be temperature constraints which will result in less energy being shifted. Second, the rebound must happen immediately after reducing power consumption. Otherwise, the temperature deviation becomes too big for too long.
The cost saving from load shifting is directly proportional to the volume and price difference between the baseline load and shifted load as given by
%
\begin{equation}\label{eq:load_shifting_savings}
    \sum_{h=1}^{24} \lambda^{\rm{s}}_{h} \big(p^{\rm{Base}}_{h} -  p_{h}\big),
\end{equation}
%
where $p_{h}$ is the power profile when the load is shifted, whereas $p^{\rm{Base}}_{h}$ is the baseline in the day-ahead stage.

>>>>

\begin{document}
\title{
    %A Pioneering Market Framework for Promoting Demand-side Flexibility in Denmark 
    %A New Ecosystem for Demand-side Flexibility Aggregators in Denmark 
    Load Shifting Versus Manual Frequency Reserve:  Which One is More Appealing \textcolor{red}{to Thermostatically Controlled Loads in Denmark}?
}
% Pioneering Market Framework for the Future Power System: the Case of Denmark
%\author{Peter Alexander Vistar Gade$^{\textrm{a}}$, Henrik Bindner, Jalal
%Kazempour}
\author[1,2]{Peter A.V. Gade\footnote{Corresponding author. Tel.: +45 24263865. \\ Email addresses: pega@dtu.dk (P.A.V. Gade), Trygve.Skjotskift@ibm.com (T. Skjøtskift), chazi@dtu.dk (C. Ziras), hwbi@dtu.dk (H.W. Bindner), jalal@dtu.dk (J. Kazempour).}}
\author[2]{Trygve Skjøtskift}
\author[1]{Charalampos Ziras}
\author[1]{Henrik W. Bindner}
\author[1]{Jalal Kazempour}
\affil[1]{Department of Wind and Energy Systems, Technical University of Denmark, Kgs. Lyngby, Denmark}
\affil[2]{IBM Client Innovation Center, Copenhagen, Denmark}
%\setcounter{Maxaffil}{0}
\renewcommand\Affilfont{\itshape\small}

@@

This paper investigates %how  thermostatically controlled loads can deliver flexibility 
demand-side flexibility provision in two  \textcolor{red}{distinct} forms of manual Frequency Restoration Reserve (mFRR) services and load shifting, and explores which one is financially more appealing to \textcolor{red}{Thermostatically Controlled Loads (TCLs)} in Denmark. While mFRR is an ancillary service required for the system and being bought by the system operator, load shifting is an individual act of the \textcolor{red}{TCL} in response to the variation of hourly electricity prices.  %We discuss advantages and disadvantages of the two and their appeal from a monetary point of view.
Without loss of generalization, we consider a supermarket freezer  as a representative \textcolor{red}{TCL}, and develop a grey-box model  describing its temperature dynamics using real data from a supermarket in Denmark. Taking into account price and activation uncertainties, a stochastic mixed-integer linear program is formulated to maximize the flexibility value from the freezer. %For practical reasons, we propose a linear policy to determine regulating power bids.
%, and then linearize the mFRR activation conditions. 
%For computational ease, we develop a decomposition technique, splitting the problem to a set of smaller sub-problems, one per scenario. %An out-of-sample evaluation was done on unseen Danish market price data for 2022. 
%
%We observed that load shifting was more profitable, but had a greater impact on the air and food temperatures in the freezer as opposed to mFRR that depends on the system state and bid price for activation.
%
%solution strategies for mFRR are presented: one with a simple policy and the other one with a dynamically updated policy. The proposed model accurately captures the timeline of decisions for bidding in manual frequency reserves.
%by using McCormick relaxation.
Examined on an ex-post simulation based on  Danish spot and balancing market prices in 2022, load shifting shows to be \textcolor{red}{almost as profitable as mFRR provision, although it could  be more} consequential for temperature deviations in the freezer. This indicates the need for regulatory measures by the Danish  system operator to make the \textcolor{red}{attraction of ancillary service provision  more obvious for TCLs in comparison to the load shifting alternative.}

@@

Power consumers are looking to reduce costs by any means due to sky-rocketing energy and gas prices. In Denmark, supermarkets are especially exposed due to their high energy consumption, and they are actively looking into initiatives to reduce their electricity bill. One of these initiatives is the application of demand-side flexibility, whereby \textcolor{red}{supermarket freezers as  Thermostatically Controlled Loads (TCLs)}  can shift consumption in time when spot (\textcolor{red}{also known as} day-ahead) prices are comparatively lower or when the power system needs up-regulation services. There are thousands of supermarkets in Denmark willing to provide flexibility, and their freezers are especially suited for this \textcolor{red}{purpose} as they constitute a big share of their energy consumption.

@@

In \cite{petersen2012eso2} and \cite{pedersen2013direct}, a complete white-box model of a supermarket refrigeration system is presented and validated against real-life data. It is also shown how such a system can provide demand response. \textcolor{red}{These two} work\textcolor{red}{s} \textcolor{red}{provide a} benchmark for any grey-box model of a supermarket refrigeration system. However, such an approach is hard to scale and requires complete knowledge of every refrigeration system. In \cite{pedersen2016improving}, a second-order model is used to model the food and air temperature in a freezer in an experimental setting. It is shown how food temperature has much slower dynamics than air temperature. In \cite{hao2014aggregate}, a simple first-order virtual battery model for a TCL is presented. This model also constitutes the starting point for modeling temperature dynamics in this work, as it has an intuitive interpretation. A similar bucket model is introduced in \cite{petersen2013taxonomy}.

@@

\textcolor{red}{In \cite{schaperow2019simulation,chanpiwat2020using, moglen2020optimal}}, residential air condition units are modeled using up- and down-regulation blocks characterizing flexibility. The blocks are obtained from grey-box models of households \cite{siemann2013performance}. The authors show how such a block formulation can be solved using exact and stochastic dynamic programming in the context of peak shaving demand for a utility in the U.S. The demand-side flexibility then functions as a hedge towards extreme electricity prices. Although such a use case has not been relevant in Denmark yet, the estimation of flexibility blocks is a novel idea as it avoids having to explicitly integrate a physical model into an optimization problem. However, a trade-off is that the curse of dimensionality quickly makes dynamic programming computationally intractable for a large portfolio of heterogeneous demand-side assets. Furthermore, it might be difficult to assure the Markov property \cite{MarkovProperty} when reformulating the problem even slightly. Flexibility blocks are also used in \cite{bobo2018offering} for an offering strategy. Here, the flexibility blocks are derived from measurements of residential appliances in the Ecogrid 2.0 project \cite{ecogrid}.

@@

In \cite{biegel2013electricity}, a portfolio of residential heat pumps is modeled using a linear first-order model. Individual units are lumped together as essentially one big, aggregated heat pump as described in \cite{biegel2013lumped}. The first-order model is a grey-box model incorporating the physics, i.e., the temperature dynamics in a household. Such an approach will be used in this work as well. The authors also consider how the portfolio of heat pumps can be used for load shifting and real-time bidding in the balancing market, but do not consider the mFRR service provision. However, they assume that all the assets can be controlled continuously in a  model predictive control  setting. Temperature deviations are minimized directly in the objective function using the integrated error of total deviations. This introduces a trade-off parameter in the objective function that must be tuned to weigh the economic value versus the temperature deviation compared to the setpoint.

\textcolor{red}{Similar studies exist in the literature albeit in different settings. For example, \cite{iria2018trading} uses a two-stage stochastic optimization for prosumers to bid into tertiary reserves, i.e., mFRR. A similar approach is taken in \cite{la2021mixed} although with a focus on scalability to thousands of assets and distribution of flexibility amongst prosumers. In \cite{nitsch2021economic}, it is shown how batteries can be used to provide secondary reserves. TCLs have also been investigated for provision of primary reserves in the Nordics \cite{paridari2020flexibility}, in a microgrid \cite{mendieta2020primary}, and in the Australian power grid \cite{attarha2020network}}.

\subsection{Our contributions and outline}
%
To the best of our knowledge, no work has \textcolor{red}{thoroughly investigated and compared the monetary value and incentives for TCLs to} provide mFRR services versus load shifting, although both have been studied extensively. This work aims to fill that gap by taking a holistic view on the incentives from \textcolor{red}{a supermarket's} perspective in Denmark. Furthermore, when stating the objective functions for mFRR, most studies have assumed a simplified market structure, and neglect the fact that the aggregator and the consumer are not necessarily the same entity. We discuss the consequences of this assumption in relation to the market structure in Denmark \cite{gade2022ecosystem}, and in the context of mFRR and load shifting.

@@

The rest of the paper is organized as follows. Section \ref{sec:monetizing_flex} describes modeling a supermarket freezer as a TCL. Thereafter, mFRR and load shifting are described in detail and their objective functions are stated. Section \ref{sec:OptimizationModel} presents the proposed stochastic optimization problem.
%Furthermore, the two solution strategies for solving the optimization problem for mFRR are introduced. 
Section \ref{sec:results} provides  results for a case study. Section \ref{sec:conclusion} concludes the paper.  \textcolor{red}{Appendix A} presents the full model formulation. Finally, \textcolor{red}{Appendix B provides a sensitivity analysis on the number of scenarios.}

\section{Describing flexibility of a TCL}\label{sec:monetizing_flex}

There are several \textcolor{red}{distinct} ways to \textcolor{red}{exploit and} monetize flexibility from a \textcolor{red}{TCL}. In this section, we focus on mFRR and load shifting. First, we describe how to mathematically model a TCL as a flexible resource. Second, we describe how to monetize the flexibility from TCLs for mFRR and load shifting, and provide the objective functions in both cases. 
For mFRR, the objective function includes all costs and revenues for the BRP, while for load shifting, the objective function only includes the flexible demand's perspective. This approach explicitly shows the situation where flexible demand is activated without including the BRP, as this is more realistic.

\subsection{Modeling TCL as a flexible resource}
%
TCLs are characterized by being controlled such that the temperature is kept at a specified setpoint. Examples include heat pumps, freezers, air condition units, etc. They constitute an important part of demand-side flexibility due to  inherent thermal inertia of such temperature-driven systems \textcolor{red}{\cite{koch2011modeling,hao2014aggregate}}.

We focus on freezers only, which are a very common type of TCLs. Specifically, we focus on a single freezer display in a Danish supermarket. Freezers are characterized by a large thermal inertia due to the frozen food, which makes them suitable for flexibility \textcolor{red}{provision}. On the other hand, there is a risk of food degradation when utilizing flexibility. Therefore, it is important to model the temperature dynamics in the freezer for a realistic and risk-aware estimation of its flexibility.

@@

\begin{figure}[t]
    \centering
    \includestandalone[width=\columnwidth]{figures/timeline_mfrr_variables_tikz}
    \caption{\textcolor{red}{Decision-making timeline and variables for the TCL when providing flexibility either in the form of mFRR services (green boxes) or load shifting (blue box).}}
    \label{fig:timeline_mfrr_variables}
\end{figure}

@@

\subsection{Load shifting}
%
Another option for utilizing flexibility is to shift the load to a different time according to the spot market prices which are known already 12-36 hours in advance \textcolor{red}{as shown in Figure \ref{fig:timeline_mfrr_variables}}. \textcolor{red}{We note that in Europe and Denmark, day-ahead prices are directly exposed to consumers with no time-of-use price in between as is often the practice in North America}. Then, it is simply a matter of consuming in low-price hours.
%
For a TCL, there are additional constraints to how energy can be shifted due to the rebound effect. First, there can be temperature constraints which will result in less energy being shifted. Second, the rebound must happen immediately after reducing power consumption. Otherwise, the temperature deviation becomes too big for too long.
The cost saving from load shifting is directly proportional to the volume and price difference between the baseline load and shifted load as given by
%
\begin{equation}\label{eq:load_shifting_savings}
    \sum_{h=1}^{24} \lambda^{\rm{s}}_{h} \big(p^{\rm{Base}}_{h} -  p_{h}\big),
\end{equation}
%
where $p_{h}$ is the power profile when the load is shifted, whereas $p^{\rm{Base}}_{h}$ is the baseline in the day-ahead stage.
